---
title: "Prediction Assignment Writeup"
author: "Shiqi Yang"
date: "9/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

## Data Loading and Exploring
```{r}
# load data
training <- read.csv('./pml-training.csv', header=T)
testing <- read.csv('./pml-testing.csv', header=T)
dim(training)
dim(testing)

# explore data
str(training)

```

We can notice that many columns have NA values or blank values on almost every observation. So we will remove them, because they will not produce any information. The first seven columns give information about the people who did the test, and also timestamps. We will not take them in our model.
```{r}
# remove columns having at least 90% of NA or blank values on the training dataset
indColToRemove <- which(colSums(is.na(training)|training=="")>0.9*dim(training)[1]) 
TrainDataClean <- training[,-indColToRemove]
# remove first 7 columns
TrainDataClean <- TrainDataClean[,-c(1:7)]
dim(TrainDataClean)

# remove columns having at least 90% of NA or blank values on the testing dataset
indColToRemove <- which(colSums(is.na(testing) |testing=="")>0.9*dim(testing)[1]) 
TestDataClean <- testing[,-indColToRemove]
dim(TestDataClean)
```

## Correlation Analysis
```{r}
library(corrplot)
corMatrix <- cor(TrainDataClean[, -53])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```


## Model Building
Three methods will be applied - Random Forests, Classification Tree, and Gradient Boosted Model. In order to limit the effects of overfitting, and improve the efficicency of the models, we will use the cross-validation technique (3 folds).

### Random Forests
```{r, cache=TRUE}
library(caret)
library(rattle)
library(rpart)

# Data partition
set.seed(12345)
inTrain1 <- createDataPartition(TrainDataClean$classe, p=0.75, list=FALSE)
Train1 <- TrainDataClean[inTrain1,]
Test1 <- TrainDataClean[-inTrain1,]
# Model fit
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFit_RandForest <- train(classe ~ ., data=Train1, method="rf", trControl=controlRF)
modFit_RandForest

# predict on test set
predict_RandForest <- predict(modFit_RandForest, newdata=Test1)
confMat_RandForest <- confusionMatrix(predict_RandForest, Test1$classe)

# display confusion matrix and model accuracy
confMat_RandForest$table; confMat_RandForest$overall[1]

plot(modFit_RandForest,main="Accuracy of Random forest model by number of predictors")

# Compute the variable importance 
MostImpVars <- varImp(modFit_RandForest)
MostImpVars
```
It appears that the optimal number of predictors is two, which provides 99% accuracy using cross-validation with 3 folds. The out of sample error is expected to be 1%. The accuracy drops as the number of predictors increases. The top two most important variables are roll_belt and yaw_belt.

### Classification Trees
```{r, cache=TRUE}
# model fit
set.seed(12345)
ControlCT <- trainControl(method="cv", number=3)
modFit_Tree <- train(classe ~ ., data=Train1, method="rpart", trControl=ControlCT)
fancyRpartPlot(modFit_Tree$finalModel)

# prediction on Test dataset
predict_Tree <- predict(modFit_Tree, newdata=Test1)
confMat_Tree <- confusionMatrix(predict_Tree, Test1$classe)

# display confusion matrix and model accuracy
confMat_Tree$table; confMat_Tree$overall[1]
```
The classification tree provides only 49% accuracy which is not a good model for predicting "classe". The out of sample error is expected to be 51%.

### Gradient boosting machine
```{r, cache=TRUE}
# model fit
set.seed(12345)
ControlCT <- trainControl(method="cv", number=3)
modFit_GBM <- train(classe~., data=Train1, method="gbm", trControl=ControlCT, verbose=FALSE)
modFit_GBM

# prediction on Test dataset
predict_GBM <- predict(modFit_GBM,newdata=Test1)
confMat_GBM <- confusionMatrix(predict_GBM, Test1$classe)

# display confusion matrix and model accuracy
confMat_GBM$table; confMat_GBM$overall[1]

```
The gradient boosting machine model provides 96% accuracy. The out of sample error is expected to be 4%.

## Conclusion
The random forest model provides the highest accuracy, which will be used to predict classe for the test data set.
```{r, cache=TRUE}
FinalPredict_RandForest <- predict(modFit_RandForest, newdata=TestDataClean)
FinalPredict_RandForest
```

